Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-03-20 paper-1/manuscript.tex
Better fig name: 4 -> threshold initial

To understand why the same stimulus A evoked a higher response in \dev{} than in \msc{} with TA only, we first compared the average responses in the two sequences, sorting neurons by their response onset to permit a high-resolution view of the stimulus response as it propagates through the network. As shown in \FIG{4}a (left and middle), we found that the response developed in a qualitatively similar fashion in both sequences. The contrast (right) revealed a clear increase in \dev{} spikes, particularly in the later parts of the response, as shown by the predominance of orange areas, which indicate higher \dev{} activity. Since the stimulus was identical in both sequences, and TA was the only plasticity mechanism in play, these differences had to be due to (1) different firing thresholds, and/or (2) different synaptic currents as a consequence of different prior activity within the trial.

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig:4}
\end{figure*}
\textbf{Fig. 4} Responses to \dev{} target trials are larger due to lower initial $\theta$.
\textbf{a} Post-stimulus spike histogram across target trials in the sample network, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and sequences. Left and middle column, target trials in the \dev{} sequence and \msc{} sequence, respectively; right column, contrast between these two.
\textbf{b} Target trial average of the threshold adaptation voltage $\theta$, with neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with neuron order and columns as in panel a. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the colormap.
\textbf{d} Relationship across excitatory neurons of the sample network between the contrast (dev - msc) in $\theta$ at the start of target trials ($\Delta \T{A}{i}$) and the contrast in the average number of spikes fired in target trials ($\Delta \R{A}{i}$), and associated linear regression.
\textbf{e} $\Delta \T{A}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \T{A}{i}$ vs $\Delta \R{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant negative correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $\theta$ and $V_m$ to bins with increased firing in \dev{} (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $\theta$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $\theta$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

We quantified the former with the strength of adaptation $\theta$ (\EQ{TA}), i.e., the amount by which the firing threshold was raised due to TA. As shown in \FIG{4}b, there was a clear and widespread lowering of thresholds at the start of target \dev{} trials relative to \msc{}, visible both in the raw data (left and middle panels) and in the contrast (right panel, early blue portions). On the other hand, once the bulk of activity had occurred, many neurons showed higher thresholds in \dev{} (right panel, later orange portions), since they spiked more often. Finally, note that inhibitory neurons, which were modeled without TA, naturally showed no difference in $\theta$, leading to numerous horizontal breaks in the plot.

To assess the difference in synaptic currents in a way that allowed direct comparison to $\theta$, we refer to the membrane voltage $V_m$, which is largely independent from and complementary to the firing threshold, and driven mostly by presynaptic inputs, resets after spikes notwithstanding. As the immediate cause of neuronal spiking, the overall pattern of $V_m$ (\FIG{4}c) largely reflected that seen in spiking activity. The contrast (right panel) shows not only a notable relative increase in membrane voltages in \dev{} around the typical response time (note the orange band running along the response front), but also an increased hyperpolarization starting immediately after due to post-spike resets.

Based on these measures, we then asked why there were more spikes in \dev{} trials. We first focused on the contribution of $\theta$ alone to confirm that thresholds are lower in \dev{}, and that this reduction in thresholds causes increased firing. We quantified the contrast in $\theta$ at the start of target trials as $\Delta \T{A}{i} = \T[dev]{A}{i} - \T[msc]{A}{i}$, where $\T{}{i} = \theta_i(t=0)$, corresponding to the leftmost datapoints of the contrast histogram in \FIG{4}b, and correlated this with the target trial response size contrast $\Delta \R{A}{i}$ across excitatory neurons. In the sample network (\FIG{4}d), we found that thresholds were clearly lower in \dev{} than \msc{} (T = 2.69e+04, p = 1.58e-92, median = -0.166 mV; one-sided Wilcoxon, n = 800 excitatory neurons), and that this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 1.6e-36; one-sided Wald test, n = 800). As shown in \FIG{4}e, this relationship held across networks and stimuli: The median of $\T{A}{i}$ among excitatory neurons was lower in \dev{} than \msc{} ($\Delta \T{A}{exc} < 0$: T = 514, p = 1.67e-16, grand median = -0.094 mV; one-sided Wilcoxon, n = 120), indicating that networks were less adapted during oddball sequences. In addition, the correlation across excitatory neurons between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$ was almost universally negative (T = 40, p = 2.68e-21, median = -0.357). This indicates that neurons with a greater decrease in $\T[dev]{A}{i}$ tended to show a greater increase in their response to target stimuli.

Yet, as we saw in \FIG{4}c, the membrane voltages also differed between the two sequences. In order to assess the relative contribution of $\theta$ to the increased \dev{} response, we chose to focus exclusively on time points with higher firing probability in the \dev{} sequence, discarding data points with equal or lower firing probability relative to the \msc{} sequence. We then weighted the remaining contrasts in $\theta$ and $V_m$ by the magnitude of the response difference and summed across neurons. The resulting time courses, shown in \FIG{4}f, show how much these two measures contributed to the increased \dev{} response. While thresholds were slightly lowered throughout the response, the membrane potential was dramatically increased. In relative terms (\FIG{4}g), $\theta$ constituted a major driver of increased firing right at the onset of the response, as well as briefly towards the end of the response. A similar pattern emerged across networks and stimuli, as shown in \FIG{4}h. Here, only the initial response was driven mainly by $\theta$, while later differences were largely explained by differences in $V_m$, leading to a relative TA contribution of less than 20\% on average. This indicates that adaptation initiated and dominated the increase in firing at the start of trials, whereas the late response was enlarged primarily as a result of increased presynaptic excitation.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig-addingSTD.pdf}
    \caption{}
    \label{fig:addingSTD}
\end{figure*}
\textbf{Fig. 7} Adding STD to the model slightly enhances the effects of TA.
\textbf{a} Reductions in mean response size $\mean{\R{A}{}}$ in each sequence as a consequence of adding STD to the model, across networks and stimuli.
\textbf{b} $\Delta \R{N}{}$ (top) and $\Delta \T{B}{}$ (bottom) among early and late neurons, respectively, across networks and stimuli, in the full model with both STD and TA. Compare to \FIG{early-adaptation}c-d.
\textbf{c} $\Delta \R{N}{exc}$ across networks and stimuli in the full model (horizontal axis) and the ablated model with TA only (vertical axis). The red line is the identity function; points above this line represent networks with greater response reduction (\dev{} - \msc{}) in the full model than in the ablated model. The subtraction of the two quantities in the box plot below shows that the non-target response tends to be reduced more in the full model.
\textbf{d} $\Delta \R{N}{exc}$ across networks and stimuli in the full model, plotted against the corresponding Pearson coefficient of the correlation between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$ across excitatory neurons. Compare to \FIG{thresholds-follow-std}d.
\textbf{e} $\Delta \T{A}{exc}$ across networks and stimuli in the full and ablated models, arranged as in panel c. The subtraction of the two quantities in the box plot below shows that $\T{A}{}$ tends to be reduced more in the full model.
\textbf{f} $\Delta \T{A}{exc}$ across networks and stimuli in the full model, plotted against the corresponding Pearson coefficient of the correlation between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$ across excitatory neurons. Compare to \FIG{4}e.
\textbf{g} $\Delta \R{A}{}$ across networks and stimuli in the full and ablated models, arranged as in panel c. The subtraction of the two quantities in the box plot below shows that the target response tends to be increased more in the full model

@@

Finally, based on the ablated model analysis in \FIG{4}, we sought to confirm that lowered $\T{A}{}$ was responsible for the increased target response. As shown in \FIG{addingSTD}f, $\Delta \T{A}{exc}$ and $\Delta \R{A}{exc}$ were negatively correlated (T = 30, p = 2.09e-21, median = -0.268; one-sided Wilcoxon, n = 120), though the correlation was weaker in the full model (T = 1.67e+03, p = 1.49e-07). Despite the weaker correlation, however, the full model target response remained clearly increased in \dev{} over \msc{}, as shown in \FIG{addingSTD}g, and this increase was greater in the full model than in the ablated model (T = 1.94e+03, p = 0.000723), with the grand medians of $\Delta \R{A}{}$ differing by 94\%. This greater contrast, of course, directly reflects the greater DDI reported in \FIG{3}.

@@

To show that $\D{}{}$ captures not only changes in $V_m$, by definition, but also in the resulting response, we correlated $\Delta \D{A}{i}$ with $\Delta \R{A}{i}$ across neurons of the sample network. Notice that this mirrors the correlation shown in \FIG{4} between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$. As shown in \FIG{synergy}b, the relationship was clearly negative (Pearson's $\rho$ = -0.541, p = 2.28e-77; Wald test, n = 1000), indicating that higher $\D{}{}$ causes reduced responses as expected. We confirmed this relationship across networks and stimuli, see \FIG{synergy}c, and found that greater depression in \dev{} predicted a lower response in most networks (T = 724, p = 1.37e-14; one-sided Wilcoxon, n = 120). We note that calculating this same correlation in the model with STD only yields no significantly negative dependence (T = 3.99e+03, p = 0.828; see \ref{sfig-synergy}).

@@

\begin{figure*}%
    \centering
    \includegraphics{fig4-withSTD.pdf}
    \caption{}
    \label{sfig4}
\end{figure*}

>>>>

To understand why the same stimulus A evoked a higher response in \dev{} than in \msc{} with TA only, we first compared the average responses in the two sequences, sorting neurons by their response onset to permit a high-resolution view of the stimulus response as it propagates through the network. As shown in \FIG{threshold-initial}a (left and middle), we found that the response developed in a qualitatively similar fashion in both sequences. The contrast (right) revealed a clear increase in \dev{} spikes, particularly in the later parts of the response, as shown by the predominance of orange areas, which indicate higher \dev{} activity. Since the stimulus was identical in both sequences, and TA was the only plasticity mechanism in play, these differences had to be due to (1) different firing thresholds, and/or (2) different synaptic currents as a consequence of different prior activity within the trial.

\begin{figure*}%
    \centering
    \includegraphics{fig-threshold-initial.pdf}
    \caption{}
    \label{fig:threshold-initial}
\end{figure*}
\textbf{Fig. 4} Responses to \dev{} target trials are larger due to lower initial $\theta$.
\textbf{a} Post-stimulus spike histogram across target trials in the sample network, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and sequences. Left and middle column, target trials in the \dev{} sequence and \msc{} sequence, respectively; right column, contrast between these two.
\textbf{b} Target trial average of the threshold adaptation voltage $\theta$, with neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with neuron order and columns as in panel a. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the colormap.
\textbf{d} Relationship across excitatory neurons of the sample network between the contrast (dev - msc) in $\theta$ at the start of target trials ($\Delta \T{A}{i}$) and the contrast in the average number of spikes fired in target trials ($\Delta \R{A}{i}$), and associated linear regression.
\textbf{e} $\Delta \T{A}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \T{A}{i}$ vs $\Delta \R{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant negative correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $\theta$ and $V_m$ to bins with increased firing in \dev{} (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $\theta$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $\theta$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

We quantified the former with the strength of adaptation $\theta$ (\EQ{TA}), i.e., the amount by which the firing threshold was raised due to TA. As shown in \FIG{threshold-initial}b, there was a clear and widespread lowering of thresholds at the start of target \dev{} trials relative to \msc{}, visible both in the raw data (left and middle panels) and in the contrast (right panel, early blue portions). On the other hand, once the bulk of activity had occurred, many neurons showed higher thresholds in \dev{} (right panel, later orange portions), since they spiked more often. Finally, note that inhibitory neurons, which were modeled without TA, naturally showed no difference in $\theta$, leading to numerous horizontal breaks in the plot.

To assess the difference in synaptic currents in a way that allowed direct comparison to $\theta$, we refer to the membrane voltage $V_m$, which is largely independent from and complementary to the firing threshold, and driven mostly by presynaptic inputs, resets after spikes notwithstanding. As the immediate cause of neuronal spiking, the overall pattern of $V_m$ (\FIG{threshold-initial}c) largely reflected that seen in spiking activity. The contrast (right panel) shows not only a notable relative increase in membrane voltages in \dev{} around the typical response time (note the orange band running along the response front), but also an increased hyperpolarization starting immediately after due to post-spike resets.

Based on these measures, we then asked why there were more spikes in \dev{} trials. We first focused on the contribution of $\theta$ alone to confirm that thresholds are lower in \dev{}, and that this reduction in thresholds causes increased firing. We quantified the contrast in $\theta$ at the start of target trials as $\Delta \T{A}{i} = \T[dev]{A}{i} - \T[msc]{A}{i}$, where $\T{}{i} = \theta_i(t=0)$, corresponding to the leftmost datapoints of the contrast histogram in \FIG{threshold-initial}b, and correlated this with the target trial response size contrast $\Delta \R{A}{i}$ across excitatory neurons. In the sample network (\FIG{threshold-initial}d), we found that thresholds were clearly lower in \dev{} than \msc{} (T = 2.69e+04, p = 1.58e-92, median = -0.166 mV; one-sided Wilcoxon, n = 800 excitatory neurons), and that this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 1.6e-36; one-sided Wald test, n = 800). As shown in \FIG{threshold-initial}e, this relationship held across networks and stimuli: The median of $\T{A}{i}$ among excitatory neurons was lower in \dev{} than \msc{} ($\Delta \T{A}{exc} < 0$: T = 514, p = 1.67e-16, grand median = -0.094 mV; one-sided Wilcoxon, n = 120), indicating that networks were less adapted during oddball sequences. In addition, the correlation across excitatory neurons between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$ was almost universally negative (T = 40, p = 2.68e-21, median = -0.357). This indicates that neurons with a greater decrease in $\T[dev]{A}{i}$ tended to show a greater increase in their response to target stimuli.

Yet, as we saw in \FIG{threshold-initial}c, the membrane voltages also differed between the two sequences. In order to assess the relative contribution of $\theta$ to the increased \dev{} response, we chose to focus exclusively on time points with higher firing probability in the \dev{} sequence, discarding data points with equal or lower firing probability relative to the \msc{} sequence. We then weighted the remaining contrasts in $\theta$ and $V_m$ by the magnitude of the response difference and summed across neurons. The resulting time courses, shown in \FIG{threshold-initial}f, show how much these two measures contributed to the increased \dev{} response. While thresholds were slightly lowered throughout the response, the membrane potential was dramatically increased. In relative terms (\FIG{threshold-initial}g), $\theta$ constituted a major driver of increased firing right at the onset of the response, as well as briefly towards the end of the response. A similar pattern emerged across networks and stimuli, as shown in \FIG{threshold-initial}h. Here, only the initial response was driven mainly by $\theta$, while later differences were largely explained by differences in $V_m$, leading to a relative TA contribution of less than 20\% on average. This indicates that adaptation initiated and dominated the increase in firing at the start of trials, whereas the late response was enlarged primarily as a result of increased presynaptic excitation.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig-addingSTD.pdf}
    \caption{}
    \label{fig:addingSTD}
\end{figure*}
\textbf{Fig. 7} Adding STD to the model slightly enhances the effects of TA.
\textbf{a} Reductions in mean response size $\mean{\R{A}{}}$ in each sequence as a consequence of adding STD to the model, across networks and stimuli.
\textbf{b} $\Delta \R{N}{}$ (top) and $\Delta \T{B}{}$ (bottom) among early and late neurons, respectively, across networks and stimuli, in the full model with both STD and TA. Compare to \FIG{early-adaptation}c-d.
\textbf{c} $\Delta \R{N}{exc}$ across networks and stimuli in the full model (horizontal axis) and the ablated model with TA only (vertical axis). The red line is the identity function; points above this line represent networks with greater response reduction (\dev{} - \msc{}) in the full model than in the ablated model. The subtraction of the two quantities in the box plot below shows that the non-target response tends to be reduced more in the full model.
\textbf{d} $\Delta \R{N}{exc}$ across networks and stimuli in the full model, plotted against the corresponding Pearson coefficient of the correlation between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$ across excitatory neurons. Compare to \FIG{thresholds-follow-std}d.
\textbf{e} $\Delta \T{A}{exc}$ across networks and stimuli in the full and ablated models, arranged as in panel c. The subtraction of the two quantities in the box plot below shows that $\T{A}{}$ tends to be reduced more in the full model.
\textbf{f} $\Delta \T{A}{exc}$ across networks and stimuli in the full model, plotted against the corresponding Pearson coefficient of the correlation between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$ across excitatory neurons. Compare to \FIG{threshold-initial}e.
\textbf{g} $\Delta \R{A}{}$ across networks and stimuli in the full and ablated models, arranged as in panel c. The subtraction of the two quantities in the box plot below shows that the target response tends to be increased more in the full model

@@

Finally, based on the ablated model analysis in \FIG{threshold-initial}, we sought to confirm that lowered $\T{A}{}$ was responsible for the increased target response. As shown in \FIG{addingSTD}f, $\Delta \T{A}{exc}$ and $\Delta \R{A}{exc}$ were negatively correlated (T = 30, p = 2.09e-21, median = -0.268; one-sided Wilcoxon, n = 120), though the correlation was weaker in the full model (T = 1.67e+03, p = 1.49e-07). Despite the weaker correlation, however, the full model target response remained clearly increased in \dev{} over \msc{}, as shown in \FIG{addingSTD}g, and this increase was greater in the full model than in the ablated model (T = 1.94e+03, p = 0.000723), with the grand medians of $\Delta \R{A}{}$ differing by 94\%. This greater contrast, of course, directly reflects the greater DDI reported in \FIG{3}.

@@

To show that $\D{}{}$ captures not only changes in $V_m$, by definition, but also in the resulting response, we correlated $\Delta \D{A}{i}$ with $\Delta \R{A}{i}$ across neurons of the sample network. Notice that this mirrors the correlation shown in \FIG{threshold-initial} between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$. As shown in \FIG{synergy}b, the relationship was clearly negative (Pearson's $\rho$ = -0.541, p = 2.28e-77; Wald test, n = 1000), indicating that higher $\D{}{}$ causes reduced responses as expected. We confirmed this relationship across networks and stimuli, see \FIG{synergy}c, and found that greater depression in \dev{} predicted a lower response in most networks (T = 724, p = 1.37e-14; one-sided Wilcoxon, n = 120). We note that calculating this same correlation in the model with STD only yields no significantly negative dependence (T = 3.99e+03, p = 0.828; see \ref{sfig-synergy}).

@@

\begin{figure*}%
    \centering
    \includegraphics{fig-threshold-initial-withSTD.pdf}
    \caption{}
    \label{sfig4}
\end{figure*}
